---
title: "Kriging with external drift (KED)"
author: |
  Laura Delgado Bejarano  
  Agricultural Engineer, Universidad Nacional de Colombia  
  Master’s Candidate, University of Campinas (UNICAMP)
date: "2024-10-23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=T}
knitr::opts_chunk$set(echo = TRUE)
# --- Directory settings (edit these paths as needed) ---
WORK_DIR <- "G:/Mi unidad/02_Maestria/01_Projeito/" # <- edit if needed
WORK_DIR <- normalizePath(WORK_DIR, winslash = "/", mustWork = TRUE)  # standardize slashes on Windows
knitr::opts_knit$set(root.dir = WORK_DIR)  # make this the working dir for all chunks


```

# Libraries 
Installing and loading the libraries that are going to be used 

```{r}
# Package manager
if(!require(pacman)) install.packages("pacman")

# Load or install required packages
pacman::p_load(sp,gstat,ggspatial,rstudioapi,raster,
  ggplot2,gridExtra,ggstar,mapview,terra,sf,
  akima,automap,paletteer
)
```

# Cleaning the space

```{r}
rm(list = ls())   # Clear all objects
graphics.off()    # Close graphics devices
cat("\014")       # Clear console
```

# Loading the data 

## Load polygon of the area of study

```{r,poly}
poly <- sf::st_read("00_Cartobase/01_Contornos/01_Paulinia/Contorno_Paulinia.shp")

plot(poly)
poly_sf <- st_zm(poly)
```

## Create a clean and empty grid to interpolate 

```{r,grid}
r  <- raster(poly_sf, res = 5)         # base grid resolution (meters)
rp <- rasterize(poly_sf, r, 0)         # empty raster within polygon
grid <- as(rp, "SpatialPixelsDataFrame")
plot(grid)

proj4string(grid) <-CRS("+init=epsg:32723") #CRS area
proj4string(grid)

```
## Load the csv with the data
Here is important to know if the csv are separed by , or ; and change it if necessary
```{r,1cada1_Argila}
original = data.frame(read.csv(file = "02_Cenarios_amostrais/04_CSVs/Paulinia_1am_cada1ha.csv",
 header = TRUE, sep = ';'))
 head(original)
 dados = original[,c(7,8,6)] # select columns of interest (x,y,z)
 head(dados)
 dados = na.omit(dados)
 names(dados) = c( "x", "y", "Ar") # ensure names x, y, variable
 sp::coordinates(dados) = ~x+y
 sp::bubble(dados, "Ar")
 
```

#Covariates

Load covariates and resample with bilinear method in the spatial resolution wanted 
```{r}
# List auxiliary rasters (.tif/.tiff)
covariates <- list.files(
  path = "01_Covariaveis/01_Paulinia/04_Stack",
  pattern = "\\.tif(f)?$",
  full.names = TRUE)

# Use elevation raster as template if available, otherwise the first raster
elev_idx <- grep("Elevacao_", basename(covariates), ignore.case = TRUE)
template_path <- if (length(elev_idx) > 0) covariates[elev_idx[1]] else covariates[1]
reference_raster <- raster::raster(template_path)
raster::res(reference_raster) <- 5

# Read, resample and standardize rasters
all_rasters <- lapply(covariates, raster::raster)
resampled   <- raster::stack(lapply(all_rasters, function(r) raster::resample(r, reference_raster, "bilinear")))
standardized <- raster::stack(lapply(1:raster::nlayers(resampled), function(i) {
  r <- resampled[[i]]
  v <- raster::getValues(r); ok <- is.finite(v)
  v[ok] <- as.vector(scale(v[ok], center = TRUE, scale = TRUE))
  r[] <- v; r
}))

names(standardized) <- c("CEa","Elev","HSI","NDCI","SBI")
plot(standardized)

```

##Extraction 

Extraction of auxiliary information co-located with soil sampling points for data

```{r}
Values<-raster::extract(standardized,dados)
head(Values)
dados@data<-cbind(dados@data,Values)
head(dados)

#Combine with interpolation grid 
grid_cov <- raster::extract(standardized, grid)
grid@data  <- cbind(grid@data, grid_cov) #the grid is going to have the value of the covariates to interpolate

```
# Geostatistics 
Adjust the experimental semivariogram, here is necessary to define the initial values based on the experimental semivariogram:  vgm(psill (c), model, range (a), nugget (co))

The formula is going to have Z~Cov1+Cov2...+Covn
```{r,semivar_Ar_1cada2}
# Build gstat object
# Here is important to change to the variable that is going to be interpolated 
g = gstat(formula = Ar ~ CEa+Elev+HSI+NDCI+SBI, data=dados) 

# Distance helpers
print(max(dist(dados@coords))/2) #Cutoff - half of the max distance 
print(min(dist(dados@coords))) #Min distance to define the lags (width)
# Experimental semivariogram
var_exp = gstat::variogram(g, cutoff=905.09, width=80, cressie=F) 
plot(var_exp)

var(dados$Ar) # Data variance (should be close to variogram sill)
#sph
fit.sph = fit.variogram(var_exp, vgm(2000, "Sph", 200, 2000)) 
plot(var_exp, fit.sph)

#exp 
fit.exp = fit.variogram(var_exp, vgm(2000, "Exp", 200, 2000)) 
plot(var_exp, fit.exp)

#gau
fit.gauss = fit.variogram(var_exp, vgm(2500, "Gau", 200,1500)) 
plot(var_exp, fit.gauss) 

```

##LOOCV 

Cross validation to select the best fit model to krige

```{r}
# Spherical
xvalid.sph <- gstat::krige.cv(Ar ~ CEa+Elev+HSI+NDCI+SBI, locations = dados, model = fit.sph)
lm_sph     <- lm(xvalid.sph$var1.pred ~ xvalid.sph$observed)
r2_sph     <- summary(lm_sph)$r.squared
rmse_sph   <- hydroGOF::rmse(xvalid.sph$var1.pred, xvalid.sph$observed)
slope_sph  <- lm_sph$coefficients[2]
ccc_sph    <- as.numeric(epiR::epi.ccc(xvalid.sph$var1.pred, xvalid.sph$observed)$rho.c["est"])

# Exponential 
xvalid.exp <- gstat::krige.cv(Ar ~ CEa+Elev+HSI+NDCI+SBI, locations = dados, model = fit.exp)
lm_exp     <- lm(xvalid.exp$var1.pred ~ xvalid.exp$observed)
r2_exp     <- summary(lm_exp)$r.squared
rmse_exp   <- hydroGOF::rmse(xvalid.exp$var1.pred, xvalid.exp$observed)
slope_exp  <- lm_exp$coefficients[2]
ccc_exp    <- as.numeric(epiR::epi.ccc(xvalid.exp$var1.pred, xvalid.exp$observed)$rho.c["est"])

# Gaussian
xvalid.gau <- gstat::krige.cv(Ar ~ CEa+Elev+HSI+NDCI+SBI, locations = dados, model = fit.gauss)
lm_gau     <- lm(xvalid.gau$var1.pred ~ xvalid.gau$observed)
r2_gau     <- summary(lm_gau)$r.squared
rmse_gau   <- hydroGOF::rmse(xvalid.gau$var1.pred, xvalid.gau$observed)
slope_gau  <- lm_gau$coefficients[2]
ccc_gau    <- as.numeric(epiR::epi.ccc(xvalid.gau$var1.pred, xvalid.gau$observed)$rho.c["est"])


```

Creating the metrics table 

```{r,cv table }
df.r2    <- data.frame(r2_sph, r2_exp, r2_gau)
df.rmse  <- data.frame(rmse_sph, rmse_exp, rmse_gau)
df.slope <- data.frame(slope_sph, slope_exp, slope_gau)
df.ccc   <- data.frame(ccc_sph, ccc_exp, ccc_gau)


temp <- data.frame(cbind(t(df.r2), t(df.rmse), t(df.slope), t(df.ccc)))
colnames(temp) <- c("R2", "RMSE", "Slope", "LCCC")

rnames <- gsub("r2_", "", rownames(temp))
rownames(temp) <- rnames

print(temp)
```

Taking theorical parameters - it has to change for the best model

```{r}
# Extract nugget, partial sill, total sill, and range
nugget        <- fit.sph$psill[1]
partial_sill  <- fit.sph$psill[2]
sill          <- partial_sill + nugget
range         <- fit.sph$range[2]

# Spatial Dependence Index (SDI) as percentage of partial sill over total sill
SDI <- (partial_sill / sill) * 100

# Classify SDI into categories
Class <- ifelse(SDI < 20, "Very low",
         ifelse(SDI < 40, "Low",
         ifelse(SDI < 60, "Medium",
         ifelse(SDI < 80, "High", "Very high"))))

# Build summary table (note: mixing numeric and character will coerce to character)
table_params <- data.frame(
  Parameter = c("Nugget", "Sill", "Range", "SDI%", "Class"),
  Value     = c(nugget, sill, range, SDI, Class),
  stringsAsFactors = FALSE
)

table_params
```

##Krige

Doing the kriging 

```{r,ok}
# Kriging with the selected model
proj4string(dados) <- CRS("+init=epsg:32723")
proj4string(dados)

mapa = krige(Ar ~ CEa+Elev+HSI+NDCI+SBI, locations = dados, newdata=grid,model = fit.sph)
plot(mapa)
```

Exporting the raster

```{r,export}
mapaRaster <- raster(mapa)

filename<-"G:/Mi unidad/02_Maestria/01_Projeito/13_GitHub/Ar_KED_1cada1_Paulinia.tiff"
writeRaster(mapaRaster, filename , format = 'GTiff', overwrite = T)

```

Graphic with ggplot

```{r,plotting,fig.width=5, fig.height=4}
ggplot() + 
  geom_raster(data = as.data.frame(mapa), aes(fill = var1.pred, x = x, y = y)) +
  scale_fill_paletteer_c("ggthemes::Red-Gold") +
  labs(fill = expression('Argila (g*kg'^-1*')'))+
  annotation_north_arrow(which_north = "grid",height = unit(1, "cm"),
                         width = unit(0.9, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(5, "cm"),
                         style=north_arrow_fancy_orienteering())+
  annotation_scale(location = "bl", width_hint = 0.2)+
  theme_bw()
```