---
title: "Ordinary Kriging with REML"
author: "Laura Delgado Bejarano"
date: "2024-10-22"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# --- Directory settings (edit these paths as needed) ---
WORK_DIR <- "G:/Mi unidad/02_Maestria/01_Projeito"  # <- edit if needed
WORK_DIR <- normalizePath(WORK_DIR, winslash = "/", mustWork = TRUE)  # standardize slashes on Windows
knitr::opts_knit$set(root.dir = WORK_DIR)  # make this the working dir for all chunks
```

# Cleaning the space

```{r,cleaning}
rm(list = ls())   # Clear all objects
graphics.off()    # Close graphics devices
cat("\014")       # Clear console

```

# Libraries 
Installing and loading the libraries that are going to be used 

```{r,libraries,echo=FALSE}
# Libraries: install if missing, then load
pkgs <- c(
  "sp", "gstat", "ggspatial", "raster", "ggplot2", "gridExtra",
  "ggstar", "mapview", "terra", "sf", "akima", "automap",
  "paletteer", "viridis", "hydroGOF","geoR"
)

# Install any missing packages (first run on a new machine)
to_install <- pkgs[!(pkgs %in% rownames(installed.packages()))]
if (length(to_install)) {
  install.packages(to_install, repos = "https://cloud.r-project.org")
}

# Load all required packages, suppressing startup messages
invisible(lapply(pkgs, function(p) {
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}))

```

# Loading the data 

## Load polygon of the area of study

```{r,poly}
poly <- st_read("00_Cartobase/01_Contornos/01_Paulinia/Contorno_Paulinia.shp") 
plot(poly)
poly_sf <- st_zm(poly)
```

## Create a clean and empty grid to interpolate 

```{r,grid}
# Create base raster at 5 m resolution and convert to grid
r  <- raster(poly_sf, res = 5)         # base grid resolution (meters)
rp <- rasterize(poly_sf, r, 0)         # empty raster within polygon
grid <- as(rp, "SpatialPixelsDataFrame")
plot(grid)

proj4string(grid) <-CRS("+init=epsg:32723")
proj4string(grid)

```

## Load the csv with the data
Here is important to know if the csv are separed by , or ; and change it if necessary


```{r,Data}
#cargas los datos
original = data.frame(read.csv(file = "02_Cenarios_amostrais/04_CSVs/Paulinia_0.2am_cada1ha.csv",
                               header = TRUE, sep = ';'))
head(original)
dados = original[,c(7,8,6)] # select columns of interest (x,y,z)
head(dados)
dados = na.omit(dados)
names(dados) = c( "x", "y", "Ar") # ensure names x, y, variable 
sp::coordinates(dados) = ~x+y 
sp::bubble(dados, "Ar") 
```
## Geostatistics 

Adjust the experimental semivariogram - first using MoM to get the initial values to use geoR in the REML model for 50 - 100 samples (Kerry and Oliver, 2007)

```{r,semivar_mom}
# Experimental variogram (MoM) and geodata prep
Variable<-"Ar"
data_geo <- as.geodata(dados, data.col = which(names(dados) == Variable))

# Build gstat object for the target variable
g = gstat(formula = Ar ~ 1, data=dados) #Change name of variable 
# Useful distance helpers to design variogram cutoff/lag width
C<-print(max(dist(dados@coords))/2);C #Cutoff
W<-print(min(dist(dados@coords)));W #Width
var_exp = gstat::variogram(g, cutoff=C, width=89, cressie=F) 
plot(var_exp)

# Initial MoM fits (keep if you want to compare visually)
fit.sph1   <- fit.variogram(var_exp, vgm(11800, "Sph", 300, 200))
fit.exp1   <- fit.variogram(var_exp, vgm(11800, "Exp", 300, 200))
fit.gauss1 <- fit.variogram(var_exp, vgm(11800, "Gau", 300, 200))

plot(var_exp, fit.sph1)
plot(var_exp, fit.exp1)
plot(var_exp, fit.gauss1)


# Cross-validation per model

# Spherical model CV
xvalid.sph <- gstat::krige.cv(Ar ~ 1, locations = dados, model = fit.sph1)
xvalid_df_sp <- as.data.frame(xvalid.sph)
lm_sph <- lm(xvalid.sph$var1.pred ~ xvalid.sph$observed)

# Metrics
r2_sph   <- summary(lm_sph)$r.squared         # R²
rmse_sph <- hydroGOF::rmse(xvalid.sph$var1.pred, xvalid.sph$observed)  # RMSE
slope_sph <- lm_sph$coefficients[2]           # regression slope

# Exponential model CV
xvalid.exp <- gstat::krige.cv(Ar ~ 1, locations = dados, model = fit.exp1)
xvalid_df_ex <- as.data.frame(xvalid.exp)
lm_exp <- lm(xvalid.exp$var1.pred ~ xvalid.exp$observed)

# Metrics
r2_exp   <- summary(lm_exp)$r.squared
rmse_exp <- hydroGOF::rmse(xvalid.exp$var1.pred, xvalid.exp$observed)
slope_exp <- lm_exp$coefficients[2]

# Gaussian model CV
xvalid.gau <- gstat::krige.cv(Ar ~ 1, locations = dados, model = fit.gauss1)
xvalid_df_ga <- as.data.frame(xvalid.gau)
lm_gau <- lm(xvalid.gau$var1.pred ~ xvalid.gau$observed)

# Metrics
r2_gau   <- summary(lm_gau)$r.squared
rmse_gau <- hydroGOF::rmse(xvalid.gau$var1.pred, xvalid.gau$observed)
slope_gau <- lm_gau$coefficients[2]


# Build summary table for the CV performance

df.r2    <- data.frame(r2_sph, r2_exp, r2_gau)          # R² per model
df.rmse  <- data.frame(rmse_sph, rmse_exp, rmse_gau)    # RMSE per model
df.slope <- data.frame(slope_sph, slope_exp, slope_gau) # slope per model

# Combine into a single table
temp <- data.frame(cbind(t(df.r2), t(df.rmse), t(df.slope)))
colnames(temp) <- c("R2", "RMSE", "slope")

# Clean row names (remove the 'r2_' prefix)
rnames <- gsub("r2_", "", rownames(temp))
rownames(temp) <- rnames

print(temp)
```

## REML semivariogram
Based on the best model in MoM we use the initial parameters to adjust the REML kriging - When this adjustment is done the lags are not fitted to the model
It is just a visual guide but it isnt equal to the MoM adjustment

```{r,semivar_Reml}
# ---------------------------
# Selected model (from MoM)
# ---------------------------
co    <- fit.gauss1$psill[1]      # nugget from MoM spherical fit
psill <- fit.gauss1$psill[2]      # partial sill (MoM)
sill  <- fit.gauss1 + co             # total sill
range <- fit.gauss1$range[2]      # range (MoM)

# ------------------------------------------
# REML fit (geoR) -> convert to gstat::vgm
# ------------------------------------------

# Spherical model (REML)
fit_sph <- geoR::likfit(
  data_geo,
  ini.cov.pars = c(psill, range),  # c(sigmasq, phi)
  nugget       = co,
  fix.nugget   = FALSE,
  lik.method   = "REML",
  cov.model    = "spherical"
)

fit.sph <- vgm(
  psill  = fit_sph$sigmasq,   # partial sill (sigmasq)
  model  = "Sph",             # Model: Spherical
  range  = fit_sph$phi,       # range (phi)
  nugget = fit_sph$tausq      # nugget (tausq)
)

plot(var_exp, fit.sph)

# Exponential model (REML)
fit_exp <- geoR::likfit(
  data_geo,
  ini.cov.pars = c(psill, range),  # c(sigmasq, phi)
  nugget       = co,
  fix.nugget   = FALSE,
  lik.method   = "REML",
  cov.model    = "exponential"
)

fit.exp <- vgm(
  psill  = fit_exp$sigmasq,   # partial sill
  model  = "Exp",             # Model: Exponential
  range  = fit_exp$phi,       # range
  nugget = fit_exp$tausq      # nugget
)

plot(var_exp, fit.exp)

# Gaussian model (REML)
fit_gau <- geoR::likfit(
  data_geo,
  ini.cov.pars = c(psill, range),  # c(sigmasq, phi)
  nugget       = co,
  fix.nugget   = FALSE,
  lik.method   = "REML",
  cov.model    = "gaussian"
)

fit.gauss <- vgm(
  psill  = fit_gau$sigmasq,   # partial sill
  model  = "Gau",             # Model: Gaussian
  range  = fit_gau$phi,       # range
  nugget = fit_gau$tausq      # nugget
)

plot(var_exp, fit.gauss)

```
## LOOCV
Cross validation to select the model to kriging

```{r,cv_Reml}
# -------------------------------
# Cross-validation to select model
# -------------------------------

# Spherical model CV
xvalid.sph <- gstat::krige.cv(Ar ~ 1, locations = dados, model = fit.sph)
xvalid_df_sp <- as.data.frame(xvalid.sph)
lm_sph <- lm(xvalid.sph$var1.pred ~ xvalid.sph$observed)

# Metrics
r2_sph    <- summary(lm_sph)$r.squared      # R²
rmse_sph  <- hydroGOF::rmse(xvalid.sph$var1.pred, xvalid.sph$observed)  # RMSE
slope_sph <- coef(lm_sph)[2]                # regression slope

# Exponential model CV
xvalid.exp <- gstat::krige.cv(Ar ~ 1, locations = dados, model = fit.exp)
xvalid_df_ex <- as.data.frame(xvalid.exp)
lm_exp <- lm(xvalid.exp$var1.pred ~ xvalid.exp$observed)

# Metrics
r2_exp    <- summary(lm_exp)$r.squared
rmse_exp  <- hydroGOF::rmse(xvalid.exp$var1.pred, xvalid.exp$observed)
slope_exp <- coef(lm_exp)[2]

# Gaussian model CV
xvalid.gau <- gstat::krige.cv(Ar ~ 1, locations = dados, model = fit.gauss)
xvalid_df_ga <- as.data.frame(xvalid.gau)
lm_gau <- lm(xvalid.gau$var1.pred ~ xvalid.gau$observed)

# Metrics
r2_gau    <- summary(lm_gau)$r.squared
rmse_gau  <- hydroGOF::rmse(xvalid.gau$var1.pred, xvalid.gau$observed)
slope_gau <- coef(lm_gau)[2]

# --------------------------------------------
# Build CV summary table and pick the best fit
# --------------------------------------------

cv_tbl <- data.frame(
  Model = c("Sph", "Exp", "Gau"),
  R2    = c(r2_sph, r2_exp, r2_gau),
  RMSE  = c(rmse_sph, rmse_exp, rmse_gau),
  Slope = c(slope_sph, slope_exp, slope_gau),
  stringsAsFactors = FALSE
)

# Print table sorted by RMSE (ascending = better)
cv_tbl_sorted <- cv_tbl[order(cv_tbl$RMSE), ]
print(cv_tbl_sorted, row.names = FALSE)


```


Taking theorical parameters - it has to change for the best model

```{r}
# Extract nugget, partial sill, total sill, and range
nugget        <- fit.gauss$psill[1]
partial_sill  <- fit.gauss$psill[2]
sill          <- partial_sill + nugget
range         <- fit.gauss$range[2]

# Spatial Dependence Index (SDI) as percentage of partial sill over total sill
SDI <- (partial_sill / sill) * 100

# Classify SDI into categories
Class <- ifelse(SDI < 20, "Very low",
         ifelse(SDI < 40, "Low",
         ifelse(SDI < 60, "Medium",
         ifelse(SDI < 80, "High", "Very high"))))

# Build summary table (note: mixing numeric and character will coerce to character)
table_params <- data.frame(
  Parameter = c("Nugget", "Sill", "Range", "SDI%", "Class"),
  Value     = c(nugget, sill, range, SDI, Class),
  stringsAsFactors = FALSE
)

table_params
```

# Kriging

Doing the kriging with the selected model


```{r,ok}
### Interpolacao por krigagem 
proj4string(dados) <- CRS("+init=epsg:32723")
proj4string(dados)

mapa = krige(Ar ~ 1, dados, grid, model = fit.gauss)
plot(mapa)


```
Exporting the raster

```{r,export}
mapaRaster <- raster(mapa)

filename<-'13_GitHub/Ar_OK_0.2cada1_Paulinia_REML.tiff'
writeRaster(mapaRaster, filename , format = 'GTiff', overwrite = T)
```

Graphic with ggplot

```{r,plotting,fig.width=7, fig.height=6}
ggplot() + 
  geom_raster(data = as.data.frame(mapa), aes(fill = var1.pred, x = x, y = y)) +
  scale_fill_viridis_c() +
    annotation_north_arrow(which_north = "grid",height = unit(1, "cm"),
                         width = unit(0.9, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(10, "cm"),
                         style=north_arrow_fancy_orienteering())+
  annotation_scale(location = "bl", width_hint = 0.2)+
  theme_bw()
```

